---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)

```

# visdat

# What does visdat do?

Initially inspired by [`csv-fingerprint`](https://github.com/setosa/csv-fingerprint), `vis_dat` helps you visualise a dataframe and "get a look at the data" by displaying the variable classes in a dataframe as a plot with `vis_dat`, and getting a brief look into missing data patterns `vis_miss`.

The name `visdat` was chosen as I think in the future it could be integrated with [`testdat`](https://github.com/ropensci/testdat). The idea being that first you visualise your data (`visdat`), then you run tests from `testdat` to fix them.

There are currently two main commands: `vis_dat` and `vis_miss`.

- `vis_dat` visualises a dataframe showing you what the classes of the columns are, and also displaying the missing data.

- `vis_miss` visualises just the missing data, and allows for missingness to be clustered and columns rearranged. `vis_miss` is similar to `missing.pattern.plot` from the `mi` package. Unfortunately `missing.pattern.plot` is no longer in the `mi` package (well, as of 14/02/2016).

- **new!!** `vis_guess` has a guess at what the value of each cell. So "10.1" will return "double", and `10.1` will return "double", and 01/01/01 will return "date". Keep in mind that it is a **guess** at what each cell is, so you can't trust this fully. `vis_guess` is made possible thanks to Hadley Wickham's `readr` package - thanks mate!


# How to install

```{r eval = FALSE}
# install.packages("devtools")

library(devtools)

install_github("tierneyn/visdat")
```

# Example

Let's see what's inside the dataset `airquality`

```{r vis_dat}

library(visdat)

vis_dat(airquality)

```

The classes are represented on the legend, and missing data represented by grey. 

by default, `vis_dat` sorts the columns according to the type of the data in the vectors. You can turn this off by setting `sort_type == FALSE`. This feature is better illustrated using the `example2` dataset, borrowed from csv-fingerprint.

```{r}

vis_dat(example2)


vis_dat(example2, 
        sort_type = FALSE)

```

The plot above tells us that R reads this dataset as having numeric and integer values, along with some missing data in `Ozone` and `Solar.R`.

We can explore the missing data further using `vis_miss`

```{r vis_miss}

vis_miss(airquality)

```

You can cluster the missingness by setting `cluster = TRUE`

```{r vis_miss-cluster}

vis_miss(airquality, 
         cluster = TRUE)

```

The columns can also just be arranged by columns with most missingness, by setting `sort_miss = TRUE`.

```{r}

vis_miss(airquality,
         sort_miss = TRUE)

```

`vis_guess` takes a guess at what each cell is. It's best illustrated using some messy data, which we'll make here.

```{r}

messy_vector <- c(TRUE,
                  T,
                  "TRUE",
                  "T",
                  "01/01/01",
                  "01/01/2001",
                  NA,
                  NaN,
                  "NA",
                  "Na",
                  "na",
                  "10",
                  10,
                  "10.1",
                  10.1,
                  "abc",
                  "$%TG")

messy_df <- data.frame(var1 = messy_vector,
                       var2 = sample(messy_vector),
                       var3 = sample(messy_vector))

```


```{r}

vis_guess(messy_df)

```

So here we see that there are many different kinds of data in your dataframe. As an analyst this might be a depressing finding. Compare this to `vis_dat`.

```{r}

vis_dat(messy_df)

```

Where you'd just assume your data is wierd because it's all factors - or worse, not notice that this is a problem.

At the moment `vis_guess` is very slow. Please take this into consideration when you are using it on data with more than 1000 rows. We're looking into ways of making it faster, potentially using methods from the `parallel` package, or extending the c++ code from `readr:::collectorGuess`.

# Interactivity

Thanks to Carson Sievert, you can now add some really nifty interactivity into visdat by using `plotly::ggplotly`, allowing for information to be revealed upon mouseover of a cell. The code to do this can be seen below, but is not shown as the github README doesn't support HTML interactive graphics...yet.

```{r eval = FALSE}

library(plotly)

vis_dat(example2) %>% ggplotly()

```

# Road Map

**Allowing for missings to be visualised/seen when there is a small proportion**

This problem will be approached by adding in a bar at the top of the visualisation that indicates whether there are any missings in a column. I will also add into the legend the % of overall missing/absent.

**visualising expectations**

The idea here is to pass expectations into `vis_dat` or `vis_miss`, along the lines of the `expectation` command in `assertr`. For example, you could ask `vis_dat` to identify those cells with values of -1 with something like this:

```{r eval = FALSE}

data %>% 
  expect(value == -1) %>%
  vis_dat

```


# Thank you

Thank you to Jenny Bryan, whose [tweet](https://twitter.com/JennyBryan/status/679011378414268416) got me thinking about vis_dat, and for her code contributions that remove a lot of testing errors.

Thank you to Hadley Wickham for suggesting the use of the internals of `readr` to make `vis_guess` work.

Thank you to Miles McBain for his suggestions on how to improve vis guess. This resulted in making it at least 2-3 times faster.

Thanks also to Carson Sievert for writing the code that combined plotly with visdat, and for Noam Ross for suggesting this in the first place.
