---
title: "Experimental Features of visdat"
author: "Nicholas Tierney"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Shipped with visdat are some experimental features that are still a bit temperamental - `vis_compare()`, `vis_dat_ly()`, and `vis_miss_ly`. This vignette describes these features in a bit more detail.
  
## using `vis_compare()`

Sometimes you want to see what has changed in your data. `vis_compare()` helps with that. It is currently only just barely working, so keep in mind that this is very much in its beta stages.

For the sake of simplicity, let's make some changes to `iris`, and compare this new dataset.

```{r vis-compare-iris}
library(visdat)
iris_diff <- iris
iris_diff[sample(1:150, 30),sample(1:4, 2)] <- NA

vis_compare(iris_diff, iris)

```

Here the differences are marked in blue.

If you try and compare differences when the dimensions are different, you get an ugly error.

```{r vis-compare-error, eval = FALSE}

iris_diff_2 <- iris
iris_diff_2$new_col <- iris$Sepal.Length + iris$Sepal.Width

vis_compare(iris, iris_diff_2)
#> vis_compare is still in BETA! If you have suggestions or errors, post an issue at https://github.com/njtierney/visdat/issuesthe condition has length > 1 and only the first element will be usedError: `.x` (5) and `.y` (6) are different lengths
```

In the future I'm hoping to create some way of representing the added columns and rows as an additional facet-like part of the plot, perhaps using something from [gridExtra](https://cran.r-project.org/web/packages/gridExtra/vignettes/arrangeGrob.html).



