---
title: "Using visdat"
author: "Nicholas Tierney"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using visdat}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo = FALSE, include = FALSE}

knitr::opts_chunk$set(fig.width = 5,
                      fig.height = 4)

```

When you get a new data set, you need to look at the data to get a sense of what it contains and potential problems with it. That's a key phrase here "looking at the data" - what does that mean?

On the one hand, you can look at the head of the data:

```{r head-iris}

head(iris)

```

Or you can have a `glimpse` at it through `dplyr::glimpse`

```{r glimpse}

dplyr::glimpse(iris)

```

Here we see that we have doubles, and a factor. We get some insight into the data.

But we don't always have data like the canonical iris dataset. let's take a look at some data that might be a bit more typical of "messy" data.

```{r visdat-glimpse}
library(visdat)
dplyr::glimpse(typical_data)

```

Looking at this, you might then ask:

> Isn't it odd that Income is a factor? And Age is a character? 

And you might start to wonder what else is different, what else changed? 

And it might be a bit unclear where to go from there. Do you plot the data? Why does my plot look weird? What are these other strange features in the data? The `visdat` package provides visualisations of an entire dataframe at once. Initially inspired by [`csv-fingerprint`](https://github.com/setosa/csv-fingerprint), `visdat` provides tools to create heatmap-like visualisations of an entire dataframe. `visdat` provides 2 main functions: `vis_dat` and `vis_miss`.

`vis_dat()` helps explore the data class structure and missingness:

```{r load-data}

vis_dat(typical_data)

```

And the `vis_miss` function provides a custom plot for missing data.

```{r example-vis-miss}

vis_miss(typical_data)

```

The name `visdat` was chosen as it borrows from the idea of [`testdat`](https://github.com/ropensci/testdat), which provides unit testing for your data.  In a similar way, `visdat` provides visual tests, the idea being that first you visualise your data (`visdat`), then you run tests from `testdat`, or a package like `assertr`, to fix these errors.

## Using `vis_dat`

Let's see what's inside the dataset `airquality`, which contains information about daily air quality measurements in New York from May to September 1973. More information about the dataset can be found with `?airquality`.

```{r vis_dat}

library(visdat)

vis_dat(airquality)

```
The plot above tells us that R reads this dataset as having numeric and integer values, with some missing data in `Ozone` and `Solar.R`. The classes are represented on the legend, and missing data represented by grey. The column/variable names are listed on the x axis. 

By default, `vis_dat` sorts the columns according to the type of the data in the vectors. You can turn this off by setting `sort_type = FALSE`. This feature is better illustrated using the `typical_data` dataset, created using [wakefield](github.com/trinker/wakefield) and contained within `visdat`.

```{r visdat-typical}

vis_dat(typical_data)

vis_dat(typical_data, 
        sort_type = FALSE)

```

## Using `vis_miss`

We can explore the missing data further using `vis_miss`.

```{r vis_miss}

vis_miss(airquality)

```

Notice that the percentages of missingness are provided in the data. These are accurate to 1 decimal place. `vis_miss` indicates when there is a very small amount of missing data at <0.1% missingness.


```{r vismiss-new-data}

df_test <- data.frame(x1 = 1:10000,
                      x2 = rep("A", 10000),
                      x3 = c(rep(1L, 9999), NA))

vis_miss(df_test)

```

`vis_miss` will also indicate when there is no missing data at all. 

```{r vismiss-mtcars}

df_test <- data.frame(x1 = 1:10000,
                      x2 = rep("tidy", 10000),
                      x3 = rep("data", 10000))

vis_miss(df_test)

```

Columns can be arranged by columns with most missingness, by setting `sort_miss = TRUE`.

```{r vismiss}

vis_miss(airquality,
         sort_miss = TRUE)

```

And missingness can be clustered by setting `cluster = TRUE`

```{r vis_miss-cluster}

vis_miss(airquality, 
         cluster = TRUE)

```

To further explore the missingness structure in a dataset, I recommend the [`naniar`](https://github.com/njtierney/naniar) package, which provides more general tools for graphical and numerical exploration of missing values.

## Using `vis_expect()`

If you want to visualise other thigns in your data aside from missing data and the classes you can visualise expectations

## Using `vis_cor()`

## Using `vis_compare()`

Sometimes you want to see what has changed in your data. `vis_compare()` displays the differences in two dataframes of the same size. Let's look at an example.

Let's make some changes to the `chickwts`, and compare this new dataset.

```{r vis-compare-iris}
chickwts_diff <- chickwts
chickwts_diff[sample(1:nrow(chickwts), 30),sample(1:ncol(chickwts), 2)] <- NA

vis_compare(chickwts_diff, chickwts)

```

Here the differences are marked in blue.

If you try and compare differences when the dimensions are different, you get an ugly error.

```{r vis-compare-error, eval = FALSE}

chickwts_diff_2 <- chickwts
chickwts_diff_2$new_col <- chickwts_diff_2$weight*2

vis_compare(chickwts, chickwts_diff_2)
# Error in vis_compare(chickwts, chickwts_diff_2) : 
#   Dimensions of df1 and df2 are not the same. vis_compare requires dataframes of identical dimensions.
```

## Using `vis_guess()`

`vis_guess()` takes a guess at what each cell is. It's best illustrated using some messy data, which we'll make here.

```{r create-messy-vec}

messy_vector <- c(TRUE,
                  T,
                  "TRUE",
                  "T",
                  "01/01/01",
                  "01/01/2001",
                  NA,
                  NaN,
                  "NA",
                  "Na",
                  "na",
                  "10",
                  10,
                  "10.1",
                  10.1,
                  "abc",
                  "$%TG")

set.seed(1114)
messy_df <- data.frame(var1 = messy_vector,
                       var2 = sample(messy_vector),
                       var3 = sample(messy_vector))

```


```{r vis-guess-messy-df, fig.show='hold', out.width='50%'}
library(visdat)
vis_guess(messy_df)
vis_dat(messy_df)

```

So here we see that there are many different kinds of data in your dataframe. As an analyst this might be a depressing finding. We can see this comparison above.

Here, you might just assume your data is weird because it's all factors - or worse, not notice that this is a problem.

At the moment `vis_guess` is very slow. Please take this into consideration when you are using it on data with more than 1000 rows. We're looking into ways of making it faster, potentially using methods from the `parallel` package, or extending the c++ code from `readr:::collectorGuess`.

## Interactive plots (vis_*_ly)

Thanks to Carson Sievert, you can now add some really nifty interactivity into visdat plots - these will eventually by using `plotly::ggplotly`, allowing for information to be revealed upon mouseover of a cell.

```{r plotly-example}

library(plotly)

vis_dat(airquality) %>% ggplotly()
vis_miss(airquality) %>% ggplotly()

```

There are also interactive versions of most of these plots, 


`vis_dat_ly(airquality)` is shorthand to call `ggplotly(vis_dat(airquality))`. When hovering the cursor over a cell, it will display information about each each cell, the type of data, the value, the row number, and the variable.

At the moment the current implementation can be slow, as `plotly` has to work hard to translate the ggplot calls into plotly.

```{r vis-dat-ly, eval = FALSE}
vis_dat_ly(airquality)
```

## vis_miss_ly

`vis_miss_ly(airquality)` is similar to `vis_dat_ly`, a shorthand call to `ggplotly(vis_miss(airquality))` with one important difference: it is written in pure plotly, so it is very fast!

When hovering the cursor over a cell, it will display information about each each cell, the type of data, the number of missings in that case, and the number of missings in that variable.

```{r vis-miss-ly}

vis_miss_ly(airquality)

```

## vis_guess_ly

This is again similar to the other `vis_*_ly()` family calls, requiring 

```{r vis-guess-ly, eval = FALSE}
vis_guess_ly(typical_data)
```

## vis_cor_ly

## vis_expect_ly

# Future work

Future work from here is focussed on making `visdat` more stable, improving the 
speed of plotting, the interactive functions

