---
title: "Using visdat"
author: "Nicholas Tierney"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo = FALSE}

knitr::opts_chunk$set(fig.width = 5,
                      fig.height = 4)

```

When you get a new data set, you need to look at the data to get a sense of what it contains and potential problems with it. That's a key phrase here "looking at the data" - what does that mean?

On the one hand, you can look at the head of the data:

```{r head-iris}

head(iris)

```

Or you can have a `glimpse` at it through `dplyr::glimpse`

```{r glimpse}

dplyr::glimpse(iris)

```

Here we see that we have doubles, and a factor. We get some insight into the data.

But we don't always have data like the canonical iris dataset. let's take a look at some data that might be a bit more typical of "messy" data.

```{r visdat-glimpse}
library(visdat)
dplyr::glimpse(typical_data)

```

Looking at this, you might then ask:

> Isn't it odd that Income is a factor? And Age is a character? 

And you might start to wonder what else is different, what else changed? 

And it might be a bit unclear where to go from there. Do you plot the data? Why does my plot look weird? What are these other strange features in the data? The `visdat` package provides visualisations of an entire dataframe at once. Initially inspired by [`csv-fingerprint`](https://github.com/setosa/csv-fingerprint), `visdat` provides tools to create heatmap-like visualisations of an entire dataframe. `visdat` provides 2 main functions: `vis_dat` and `vis_miss`.

`vis_dat()` helps explore the data class structure and missingness:

```{r load-data}

vis_dat(typical_data)

```

And the `vis_miss` function provides a custom plot for missing data.

```{r example-vis-miss}

vis_miss(typical_data)

```

The name `visdat` was chosen as it borrows from the idea of [`testdat`](https://github.com/ropensci/testdat), which provides unit testing for your data.  In a similar way, `visdat` provides visual tests, the idea being that first you visualise your data (`visdat`), then you run tests from `testdat`, or a package like `assertr`, to fix these errors.

## Using `vis_dat`

Let's see what's inside the dataset `airquality`, which contains information about daily air quality measurements in New York from May to September 1973. More information about the dataset can be found with `?airquality`.

```{r vis_dat}

library(visdat)

vis_dat(airquality)

```
The plot above tells us that R reads this dataset as having numeric and integer values, with some missing data in `Ozone` and `Solar.R`. The classes are represented on the legend, and missing data represented by grey. The column/variable names are listed on the x axis. 

By default, `vis_dat` sorts the columns according to the type of the data in the vectors. You can turn this off by setting `sort_type = FALSE`. This feature is better illustrated using the `typical_data` dataset, created using [wakefield](github.com/trinker/wakefield) and contained within `visdat`.

```{r visdat-typical}

vis_dat(typical_data)

vis_dat(typical_data, 
        sort_type = FALSE)

```

## Using `vis_miss`

We can explore the missing data further using `vis_miss`.

```{r vis_miss}

vis_miss(airquality)

```

Notice that the percentages of missingness are provided in the data. These are accurate to 1 decimal place. `vis_miss` indicates when there is a very small amount of missing data at <0.1% missingness.


```{r vismiss-new-data}

df_test <- data.frame(x1 = 1:10000,
                      x2 = rep("A", 10000),
                      x3 = c(rep(1L, 9999), NA))

vis_miss(df_test)

```

`vis_miss` will also indicate when there is no missing data at all. 

```{r vismiss-mtcars}

df_test <- data.frame(x1 = 1:10000,
                      x2 = rep("tidy", 10000),
                      x3 = rep("data", 10000))

vis_miss(df_test)

```

Columns can be arranged by columns with most missingness, by setting `sort_miss = TRUE`.

```{r vismiss}

vis_miss(airquality,
         sort_miss = TRUE)

```

And missingness can be clustered by setting `cluster = TRUE`

```{r vis_miss-cluster}

vis_miss(airquality, 
         cluster = TRUE)

```

To further explore the missingness structure in a dataset, I recommend the [`naniar`](https://github.com/njtierney/naniar) package, which provides more general tools for graphical and numerical exploration of missing values.

## Interactivity

Thanks to Carson Sievert, you can now add some really nifty interactivity into visdat by using `plotly::ggplotly`, allowing for information to be revealed upon mouseover of a cell.

```{r plotly-example}

library(plotly)

vis_dat(airquality) %>% ggplotly()
vis_miss(airquality) %>% ggplotly()

```

# Future work

There are some other features in development for interactivity, guessing types of data, and comparing datasets. You read more about these in the vignette, ["Experimental Features of visdat"](http://www.njtierney.com/visdat/articles/experimental_features.html).

